Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-10-09 pkg/glmertree/paper/treatment_interactions.Rnw
various improvements -> still some open READMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@997 255f4249-2263-4417-b1ea-8498779e27a7

@@

@@

\begin{figure}[t!]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all $x$-variables interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. $\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above. However, the two subgroups in the middle do not represent true subgroups, which may be due to the clustered structure of the dataset not being taken into account.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$th element representing the random intercept for cluster $m$; again, for simplicity we assume that only cluster-specific intercepts are included in the model, but multiple random-effects coefficients can easily be included. Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} and random effects are incorporated as well: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take this approach a step further: instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree. This allows not only for detection of differences in main effects, but also for detection of differences in regression effects (e.g., of treatment type) across terminal nodes. In addition, GLMM trees can be estimated for continuous, as well as binary and count variables. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:

\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the GLM tree and coefficients $\beta_{j(r)}$ and $b_{(r)}$ are re-estimated. The GLM tree is estimated, given the estimated random effects from the last iteration, and the random effects are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}.

@@

\section{Simulation-based Evaluation: Method}

To assess the performance of GLMM tree, we performed three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

R \cite{R14} was used for the generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions. For other response distributions, the \verb|glmtree| function would be available. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

@@

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(3L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{enumerate}
	\item a main effect of $U_2$ (coefficient = 1.5) 
	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = -.25), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = .25)
	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = -.25), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = .25)
\end{enumerate}

@@

\section{Simulation-based Evaluation: Results}

@@

\begin{figure}[t!]
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[t!]
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ correlation.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=correlation.ind, type="b", 
ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Tree accuracy of GLM and GLMM tree. Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent levels of sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[t!]
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
scales=list(y=list(limits = c(0,1))))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Rows represent the levels of the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences, columns represent levels of sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

@@

\begin{figure}[t!]
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("noint_treesizes_long.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{figure}[t!]
<<accuracy-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.acc <- aggregate(formula=truevars ~ truevars.ind + N + type + np, FUN=mean, 
data=treespecs.long)
xyplot(truevars ~ type | N + np, data = aggdata.acc, groups=truevars.ind, type="b",
ylab="proportion of accurately recovered models", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Model accuracy for GLMMs and GLMM trees. In an accurate model, the true main and interaction effects were accurately recovered. Columns represent sample size, rows represent the number of covariates in the model.}
\label{fig:xyplot_accuracy_pw}
\end{figure}

@@

\begin{figure}[t!]
<<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + np + type + N, FUN=mean, 
data=treespecs.long)
xyplot(correlation.values ~ type | N + np, data = aggdata.cor, groups=correlation.ind, type="b",
ylab="correlation", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Average predictive accuracy of GLMMs and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Columns represent sample size, rows represent the number of covariates.}
\label{fig:xyplot_correlation_pw}
\end{figure}

@@

To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from a individual-patient data meta-analysis of \citeA{CuijyWeit14}. This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender predicted or moderated treatment outcomes. Further details on the dataset can be found in \citeA{CuijyWeit14}.

@@

The results of recursive partitioning techniques are known to be potentially unstable, in the sense that small changes in the dataset may substantially alter the variables or values selected for partitioning. Therefore, we used the R-package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values. Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning, over 500 subsamples of size $.9 \times N$ of the dataset.

@@

Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset. Note that the selection frequencies do not add up to 1, as trees may involve multiple, or no splits. Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority ofGLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples. As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.

>>>>

%% for comments
\newcommand{\readme}[1]{\emph{\marginpar{README} (#1)}}

@@

\begin{figure}[t!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a global GLM for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ yields a $p$-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). This partition is created using $U_{{k^*}}$, the variable with the minimal $p$-value in step (2). The split point for $U_{{k^*}}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected (or the subsets become too small).

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

Due to the binary recursive nature of MOB, the resulting partition can be represented as a binary tree. If the partitioning is based on a GLM, the result is a GLM tree, with a local fixed-effects regression model in every $j$-th ($j = 1,\dots,J$) terminal node or subgroup:
%
\begin{equation}
	\label{eq:fixedeffects_MOB}
	g(\mu_{ij}) = x_{i}^{\top}\beta_{j}
\end{equation}
%
Note that, if the recursive subgroup structure (i.e., the partition) were known, the tree could be estimated as a single GLM where all $x$-variables interact with the subgroup indicator. Somewhat more formally, the model could then be written: $g(\mu_{i}) = x_{i}^{* \top}\beta^{*}$, where $x_{i}^{*}$ are the values of the $2J$ interactions between the subgroups from the tree, and the elements of $x_{i}$. The corresponding $\beta^{*}$ would have length $2J$ as well, containing subgroup-specific fixed-effects coefficients.

Figure~\ref{fig:example_mobtree} shows the GLM tree grown on the motivating dataset. By using the three auxiliary covariates (anxiety, duration and age), MOB partitioned the observations into four subgroups, each with a different estimate for $\beta_j$. Age was correctly not detected as a partitioning variable, and the left- and rightmost subgroups are in accordance with the true treatment-subgroup interactions described above. However, the two subgroups in the middle do not represent true subgroups, which is due to the clustered structure of the dataset not being taken into account.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$. For simplicity, we just employ such a cluster-specific intercept here but further random effects can be easily included in $z_i$. Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

Although the random-effects part of the GLMM in Equation \ref{eq:mixedeffects} accounts for the nested structure of the dataset, the global fixed-effects part may not describe the data well. Therefore, we propose the GLMM tree model, in which the fixed-effects part may be partitioned as in Equation \ref{eq:fixedeffects_MOB} while still adjusting for random effects: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression tree (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART tree with constant fits in the nodes, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the CART tree, and (2) assuming the CART tree known, allowing for estimation of the random effects.

For estimating GLMM trees, we take this approach two steps further: (1)~Instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree. This allows not only for detection of differences in intercepts/levels across terminal nodes, but also for detection of differences in slopes/regression coefficients such as treatment effects. (2)~By using generalized linear (mixed) models, the response may also be a binary or count variable instead of a continuous variable. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the GLM tree and coefficients $\beta_{j(r)}$ and $b_{(r)}$ are re-estimated. The GLM tree is estimated, given the estimated random effects from the last iteration, and the random effects are estimated, given the estimated GLM tree from the current iteration. Iterations are continued until convergence, which is monitored by computing the log-likelihood criterion of the mixed-effects model in Equation~\ref{eq:glimmertree}. Typically, this converges if the tree does not change from one iteration to the next.

@@

\section{Simulation-based evaluation: Method}

To assess the performance of GLMM tree, we carried out three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

R \cite{R} was used for the generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| is available for other response distributions. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
%
For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

@@

\readme{Z: My impression is that the description of the data-generating process is too far ahead of the simulation itself. I would recommend to first explain what is the same in all three simulation studies. And then what is specific could maybe be included at the beginning of each of the individual studies so that readers have it in mind when looking at the results. So in this section here I would just briefly explain the idea and defer all details until later.}

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(3L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0), U5 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{enumerate}
	\item a main effect of $U_2$ (coefficient = 1.5) 
	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = $-.25$), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = $.25$)
	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = $-.25$), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = $.25$)
\end{enumerate}

\readme{Z: Also include the formula for the linear predictor in this setup so that this can be easily matched with the GLMM estimated.}

@@

\section{Simulation-based evaluation: Results}

@@

\begin{figure}[t!]
\centering
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

\begin{figure}[t!]
\centering
<<accuracy-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.acc <- aggregate(formula=truetree.values ~ correlation.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)
print(xyplot(truetree.values ~ sigmab | N + corUb, data = aggdata.acc, groups=correlation.ind, type="b", 
ylab="tree accuracy", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=TRUE, points=TRUE))
)
@
\caption{Tree accuracy of GLM and GLMM tree. Tree accuracy is defined as the proportion of datasets in which the true tree was accurately recovered. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$, columns represent levels of sample size.}
\label{fig:xyplot_treeaccuracy}
\end{figure}

@@

\begin{figure}[t!]
\centering
<<correlation-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + N + sigmab + treatdiff, FUN=mean, data=treespecs.long)
print(xyplot(correlation.values ~ sigmab | N + treatdiff, data = aggdata.cor, groups=correlation.ind, 
type="b", ylab="correlation", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T),
scales=list(y=list(limits = c(0,1))))
)
@
\caption{Average predictive accuracy of GLM and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Rows represent the levels of the absolute value of the unstandardized treatment-effect difference in subgroups with treatment-effect differences, columns represent levels of sample size.}
\label{fig:xyplot_correlations}
\end{figure}

@@

\readme{Z: Wouldn't it be more natural to just look at P(tree size greater than 1), i.e., the probability of splitting at all. Then it could be worked out more clearly that GLMM trees approximately keep the significance level while GLM trees generally have an inflated irror.}

@@

\begin{figure}[t!]
\centering
<<treesize-nointeract-xyplot, fig=TRUE, width=9, height=7>>=
load("noint_treesizes_long.dat")
treesizes.long$sigmabm <- factor(as.numeric(as.character(treesizes.long$sigmabm)), ordered = TRUE)
treesizes.long$N <- factor(as.numeric(as.character(treesizes.long$N)), ordered = TRUE)
aggdata <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmabm + corUbm, FUN=mean, data=treesizes.long)
print(xyplot(treesize.values ~ sigmabm | N + corUbm, data = aggdata, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=FALSE), 
abline=c(1,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1, 
lines=TRUE, points=TRUE)))
@
\caption{Average tree size of GLM and GLMM trees for datasets without treatment-subgroup interactions. Rows represent dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent sample size.}
\label{fig:xyplot_treesize_nointeract}
\end{figure}

@@

\begin{figure}[t!]
\centering
<<accuracy-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.acc <- aggregate(formula=truevars ~ truevars.ind + N + type + np, FUN=mean, 
data=treespecs.long)
xyplot(truevars ~ type | N + np, data = aggdata.acc, groups=truevars.ind, type="b",
ylab="proportion of accurately recovered models", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Model accuracy for GLMMs and GLMM trees. In an accurate model, the true main and interaction effects were accurately recovered. Columns represent sample size, rows represent the number of covariates in the model.}
\label{fig:xyplot_accuracy_pw}
\end{figure}

@@

\begin{figure}[t!]
\centering
<<correlation-pw-xyplot, fig=TRUE, width=9, height=7>>=
aggdata.cor <- aggregate(formula=correlation.values ~ correlation.ind + np + type + N, FUN=mean, 
data=treespecs.long)
xyplot(correlation.values ~ type | N + np, data = aggdata.cor, groups=correlation.ind, type="b",
ylab="correlation", xlab="interaction type", par.settings=standard.theme("pdf",color=F), 
auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
@
\caption{Average predictive accuracy of GLMMs and GLMM trees. Predictive accuracy of trees is defined as the correlation between the true and predicted differences between Treatment~1 and~2. Columns represent sample size, rows represent the number of covariates.}
\label{fig:xyplot_correlation_pw}
\end{figure}

@@

To further illustrate the use of GLM and GLMM tree, we applied both algorithms to a dataset from an individual-patient data meta-analysis of \citeA{CuijyWeit14}. This meta-analysis was based on patient-level observations from 14 RCTs, comparing the effects of psychotherapy (cognitive behavioral therapy; CBT) and pharmacotherapy (PHA) in the treatment of depression. The study of \citeA{CuijyWeit14} was aimed at establishing whether gender is a predictor or moderator of the outcomes of psychological and pharmacological treatments for depression. Treatment outcomes were assessed by means of the 17-item Hamilton Rating Scale for Depression (HAM-D; \citeNP{Hami60}). \citeA{CuijyWeit14} found no indication that gender predicted or moderated treatment outcomes. Further details on the dataset can be found in \citeA{CuijyWeit14}.

@@

The results of recursive partitioning techniques are known to be potentially unstable, in the sense that small changes in the dataset may substantially alter the variables or values selected for partitioning. Therefore, we used the R~package \verb|stablelearner| \cite{PhilyZeil16} to assess the stability of the selected splitting variables and values. Using the \verb|stabletree| function, we calculated the number of times a variable and value were selected for partitioning, over 500 subsamples of size $.9 \times N$ of the dataset.

@@

Table \ref{tab:stability} presents statistics on the variables selected for partitioning in subsamples of the dataset. Note that the selection frequencies do not add up to 1, as trees may involve multiple, or no splits. Table \ref{tab:stability} indicates that the presence of a comorbid anxiety disorder was selected for partitioning in the majority of GLMM trees grown on subsamples of the dataset, and all other variables were selected in less than 3\% of the subsamples. As the comorbid anxiety disorder variable involved only a single splitting value, further assessment of the stability of splitting values was not necessary.
